---
title: "Data 607 Assignment 1 - Loading data into a Data Frame"
author: "Lawrence Yu"
date: "`r Sys.Date()`"
output: openintro::lab_report
---


### Overview
https://fivethirtyeight.com/features/how-our-mlb-predictions-work/
This article explains how the writers settled on their current rating system for predicting a baseball team's likelihood of winning a game. Included are the different factors that are considered and the rationale behind their weightings. The ultimate goal of this system is to predict which team will make the postseason and finally win the World Series every season.

### Data Frame and Modifications
```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(RCurl)
library(dplyr)
```

```{r mlb_latest_dataset_columns }
mlb_elo_latest_url = "https://raw.githubusercontent.com/Megabuster/Data607/refs/heads/main/data/Assignment1/mlb_elo_latest.csv"
raw_text <- getURL(mlb_elo_latest_url)
mlb_elo_latest_dataset <- read.csv(text = raw_text)
mlb_elo_parsed_dataframe <- select(mlb_elo_latest_dataset, team1, team2, rating_prob1, rating_prob2)
colnames(mlb_elo_parsed_dataframe)[colnames(mlb_elo_parsed_dataframe) == "rating_prob1"] <- "rating_win_probability1"
colnames(mlb_elo_parsed_dataframe)[colnames(mlb_elo_parsed_dataframe) == "rating_prob2"] <- "rating_win_probability2"
mlb_elo_parsed_dataframe[1:10,]
```

### Findings
The article compared the Elo system and team ratings to show why FiveThirtyEight uses the latter for its seasonal predictions. The current team ratings system is very adaptive in that it considers team strength yet accounts for factors such as team rest that can affect how a team performs. This system, while better than the historic Elo system, can still be improved further.

### Recommendations
Curiously, home-field advantage is given a flat 24 rating points with an adjustment when there are no fans. I believe home-field advantage is far more nuanced than the current iteration of the model gives it credit for. The article references this Fangraphs article, https://library.fangraphs.com/the-beginners-guide-to-understanding-park-factors/, but fails to account for the difference in park factors that the latter article describes. Baseball fields do not follow clear dimensions standards. This means it is unlikely that all stadiums have a comparable home-field advantage. The field on which the game is played at is a known and visible factor. This makes it a good way to improve the model.


